---
title: "Gov 50 Final Project"
author: "Chris Ruiz"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

MILESTONE 1: 

I am interested in exploring data related to education. I volunteer at the Cambridge After School Program near Alewife and would love to see if I can collect data through surveys of the students about their thoughts on the effectiveness of the program, or really anything related to education. I don't really know how this will manifest itself, but its a good place to start and I'm excited to brainstorm!


MILESTONE 2:
The dataset that I will be using is called School Attendance by Student Group and District, 2021-2022. It is from data.gov. The dataset takes into account schools in the state of Connecticut (public and private) and also stores factors like race, homelessness, income level, whether or not students are on free-reduced lunch programs, school district, and attendance rates between 2019 and 2022. The general question I would like to answer is did COVID-19 impact the attendance rates of low income students and minority students, and if so, by how much, and also if so, did students of color have consistently lower attendance rates than their white counterparts? I hypothesize that COVID disproportionately impacted poorer students and students of color and that their attendance rates in the state of Connecticut will be lower across the board. I have read many articles that claim that COVID impacted the minority workforce disproportionately, so I believe that my hypothesis is rooted in logic and theory. Although I can't establish causation, I think that this may be true because students in lower income districts had worse access to vaccines, more exposure to the virus, and less school resources to contain the spread of the virus, in turn lowering their attendance rates compared to wealthier, whiter districts. The variables I plan to use are percentage of non white students, percentage of white students, attendance rates in 2019, attendance rates in 2020, attendance rates in 2021, as well as the other factors like homelessness, income, and need. If all goes well, I will establish a negative correlation between non white percentage and attendance rates. This will be measured with a correlation coefficient from -1 to 1. My hypothesis would be proven if the data, when plotted, showed a negative trend between minorities and high attendance rates. My hypothesis would be disproven if attendance rates between 2019 and 2021 were consistent, and also if attendance rates between non whites and whites were consistent. I am excited to get into the data and see what I can find!

MILESTONE 3: 

In this milestone, I was interested in overall attendance rates among Black and White students. I chose 2022 for this first visualization, but plan to analyze attendance impact over all 3 years that the dataset allows. It is clear that in the following graphs, black attendance rates are lower than white attendance rates across the board.

MILESTONE 4: 

By analyzing the attendance distributions below, there is evidence that Black and Latino students, over all Connecticut school districts, had lower attendance rates in the 2020-2021 school year. I plan to analyze a connection between COVID-19 and attendance rates to compare attendance rates both before and after COVID-19. However, with the data visualizations below, I can estimate that the mean attendance in 2021 for White students is a couple percentage points higher than their Latino and Black counterparts. I also plan to see how homelessness impacted attendance rates, but for right now, the data supports my hypothesis that COVID-19 is not colorblind and impacts communities of color differently. 

```{r, echo = FALSE}
library(tidyverse)
library(dplyr)

attendance <- read_csv("~/Downloads/attendance.csv", skip = 1)

new_names <- c("code", "district", "category", "group", "student", "att22", "rate22", "att21", "rate21", "att20", "rate20", "period", "update")

colnames(attendance) <- new_names

print(colnames(attendance))

attendance
```

```{r, echo = FALSE}
white_attendance_21 <- attendance |>
  filter(group == "White", category == "Race/Ethnicity") |>
  select(att21) |>
  ggplot(mapping = aes(x = att21)) +
  geom_histogram(binwidth = .005) +
  labs(x = "Attendance in 2021 School Year", y = "Count", title = "White Attendance in Connecticut Schools 20-21")

 black_attendance_21 <- attendance |>
  filter(group == "Black or African American", category == "Race/Ethnicity") |>
  select(att21) |>
  ggplot(mapping = aes(x = att21)) +
  geom_histogram(binwidth = .005) +
  labs(x = "Attendance in 2021 School Year", y = "Count", title = "Black Attendance in Connecticut Schools 20-21")
 
latino_attendance_21 <- attendance |>
  filter(group == "Hispanic/Latino of any race", category == "Race/Ethnicity") |>
  select(att21) |>
  ggplot(mapping = aes(x = att21)) +
  geom_histogram(binwidth = .005) +
  labs(x = "Attendance in 2021 School Year", y = "Count", title = "Hispanic/Latino Attendance in Connecticut Schools 20-21")
```


```{r}
white_attendance_21
black_attendance_21
latino_attendance_21
```

```{r}
